The promise of community driven preprints in ecology and evolution

### 2024-04-19 ms/ms.qmd
First draft. Off to Shin and Losia

Preprint servers, such as *EcoEvoRxiv*, can empower researchers to make their findings more accessible, open and transparent but only if they are used well and initiatives are widely known. Significant barriers to preprint use still exist globally – from lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by a research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution which can help inform future initiatives, improve others, and allow us to work harder in further breaking down barriers to their use in our field.

@@

### **Science takes time, but publication takes longer**
Increased competition in science has resulted in the bar being raised with respect to the amount of data required for publication [@vale2015accelerating]. This is a good outcome if it results in higher impact research that better clarifies our understadning of the natural world, but it does come at a cost for the rate at which research findings can be disseminated [@vale2015accelerating]. Preprinting has been proposed as a way research can be disseminated more quickly as it can take a long time before results are ultimately published [@bourne2017ten; @vale2015accelerating]. However, data on the time to publication is sparse, and the real benefit of preprints in this context would be reinforced if preprinting is able to dramatically speed up dissemination. We evaluated just how much time it takes to publish a research paper in ecology and evolution by ascertaining the time between when a preprint was first posted to when it was finally accepted in a research journal. In total, `r unpub` papers were still unpublished out of the `r nrow(data2)` (`r (unpub / nrow(data2))*100`%) at the time these data were collected. Astonishingly, the maximum time between submission and publication was `r sum_preprints$max` days ([@fig-pubsummary]A), and the median time to publication was `r sum_preprints$median` days (mean = `r sum_preprints$mean`; SD = `r sum_preprints$sd` days). Our results largely confirm the long timespans most authors are dealing with to get their research published after the paper is written.

### **Catious 'open'-mindedness of research findings in preprints**
In additon to speeding up dissemination, preprints can also be a useful way to ensure that research remains open and accessible to the research community because the paper is available online no matter what the open access status of the publication ends up being [@bourne2017ten; @vale2015accelerating]. Using the published articles on *EcoEvoRxiv*, we evaluated the open access status of each paper using....LOSIA?. Most of the published preprints ended up being open access([@fig-pubsummary]A), however, the status of many (i.e., `r sum_is_oa$n[3]` or `r sum_is_oa$percent[3]*100`%) were unknown, and `r sum_is_oa$n[1]` or `r sum_is_oa$percent[1]*100`% were not published in open access reserach journals. The level of open access for a paper published in an open access journal also varied widely ([@fig-pubsummary]B).

@@

- First, authors need not be afraid of sharing data and code early in the publication process. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings [@sarabipour2019value]. If authors are worried about data being used unintentially, clear information surrounding its re-use and withholding some data or meta-data needed to understand the dataset may be options. 
- Second, authors need to be aware of the benefits of posting preprints and postprints. Posting preprints is expected to help: 1) establish precedence of research findings; 2) ensure that research is open access and 3) authors receive feedback on their research before it is published [@bourne2017ten]. While it is clear points 1 and 2 have already been achived discussions around a preprints findings are still rare, which is consistent with findings on other preprint servers (e.g. BioRxiv [@anderson2020biorxiv]). It is clear that we, as a community, need to provide better platforms that document discussions around a preprint findings. Such discussions are useful for the authors to help improve their work and communicate their findings more effectively (when done constructively, of course).
- Third, authors need to be aware of the benefits of using community driven preprint servers and the unique initiatives being pursued. Community driven preprint servers are capable of setting their own agenda and are driven by the needs and disires of the community they would like their findings to be shared with. Supporting these initiatives should be a priority for all researchers and those volunteering their time at *EcoEvoRxiv* remain open to new and innovative ways to improve publication and open science practices. 
- Lastly, it remains clear that the time between preprinting and publication is still quite long. One possible explanation is that preprints are not being sent to journals deemed suitable for the work. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recomendations when deciding to publish a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process.

While most preprints get seemlessly connected with and merged with their published version some remain disconnected. Incorrect cross-linking can create confusion in research findings and lead to frustration from authors. Thankfully, the reasons for unmatched preprints and publications are fairly well understood and easily rectified. They often result from a mismatch between titles or authors in preprint and published metadata. In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%). Mismatched metadata was the almost always the contributer to preprints and published articles not being matched automatically in Google Scholar during our data collection. As such, we would recommend that authors make sure that, when preprints end up getting published, that they update their preprint with the publication DOI when it is available. Particularly if their title has changed. This is quite easy for authors to do in *EcoEvoRxiv* and it will ensure that the preprint is linked to the published article and the citations are merged. It will also create a link between the final preprint and the published paper so that, no matter the open access status of the publication, the reserach findings will remain openly accessible, saving costs for authors and improving the visibility and use of research.

Despite the early successes of the new initiatives taken by *EcoEvoRxiv* it is still clear that much work needs to be done to improve our communities understanding of preprints (and postprints). We view this perspective piece as one small step in achieving that goal. We believe that the future of preprints is bright and community driven initiatives, such as *EcoEvoRxiv*, are set to play a key role in the future of scientific publishing.

>>>>

Preprint servers, such as *EcoEvoRxiv*, can empower researchers to make their findings more accessible, open and transparent but only if they are used well and initiatives are widely known. Significant barriers to preprint use still exist globally – from lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by a research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution which can help inform future initiatives, improve others, and allow us to work harder in further break down barriers to their use in our field.

@@

### *Science takes time, but publication could take longer*
Increased competition in science has resulted in the bar being raised with respect to the amount of data required for publication [@vale2015accelerating]. This is a good outcome if it results in higher impact research that better clarifies our understanding of the natural world, but it does come at a cost for the rate at which research findings can be disseminated [@vale2015accelerating]. Preprinting has been proposed as a way research can be disseminated more quickly as it can take a long time before results are ultimately published [@bourne2017ten; @vale2015accelerating]. However, data on the time to publication is sparse, and the real benefit of preprints in this context would be reinforced if preprinting is able to dramatically speed up dissemination. We evaluated just how much time it takes to publish a research paper in ecology and evolution by ascertaining the time between when a preprint was first posted to when it was finally accepted in a research journal. In total, `r unpub` papers were still unpublished out of the `r nrow(data2)` (`r (unpub / nrow(data2))*100`%) at the time these data were collected. Astonishingly, the maximum time between submission and publication was `r sum_preprints$max` days ([@fig-pubsummary]A), and the median time to publication was `r sum_preprints$median` days (mean = `r sum_preprints$mean`; SD = `r sum_preprints$sd` days). Our results largely confirm the long timespans most authors are dealing with to get their research published after the paper is written.

### *Catious 'open'-mindedness of research findings in preprints*
In additon to speeding up dissemination, preprints can also be a useful way to ensure that research remains open and accessible to the research community because the paper is available online no matter what the open access status of the publication ends up being [@bourne2017ten; @vale2015accelerating]. Using the published articles on *EcoEvoRxiv*, we evaluated the open access status of each paper using....LOSIA?. Most of the published preprints ended up being open access([@fig-pubsummary]A), however, the status of many (i.e., `r sum_is_oa$n[3]` or `r sum_is_oa$percent[3]*100`%) were unknown, and `r sum_is_oa$n[1]` or `r sum_is_oa$percent[1]*100`% were not published in open access reserach journals. The level of open access for a paper published in an open access journal also varied widely ([@fig-pubsummary]B).

@@

- First, authors need not be afraid of sharing data and code early in the preprint stage. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings [@sarabipour2019value]. If authors are worried about data being used unintentially, clear information surrounding its re-use and withholding some data or meta-data needed to understand the dataset may be options. 
- Second, authors need to be aware of the benefits of posting preprints and postprints for constructive feedback [@bourne2017ten]. While it is clear preprints help establish precedence and allow findings to be openly accessible it is still rare that constructive discussions form around preprints in an open form, which is consistent with findings on other preprint servers (e.g. BioRxiv [@anderson2020biorxiv]). It is clear that we, as a community, need to provide better platforms that document discussions around a preprint findings. Such discussions are useful for the authors to help improve their work and communicate their findings more effectively (when done constructively, of course). One way to facilitate such discussions may be to use open peer reviewing serves such as *Peer Community In* (PCI) or *PubPeer* to provide feedback on preprints that ultimately are associated with a preprint.
- Third, it remains clear that the time between preprinting and publication is still quite long. One possible explanation is that preprints are not being sent to journals deemed suitable for the work. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recomendations when deciding to publish a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process. 
- Finally, while most preprints get seemlessly connected with and merged with their published version some remain disconnected. Incorrect cross-linking can create confusion and lead to frustration among authors. Thankfully, the reasons for unmatched preprints and publications are fairly well understood and easily rectified. They often result from a mismatch between titles or authors in preprint and published metadata. In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%). Mismatched metadata was almost always the main contributer to preprints and published articles not being matched automatically in Google Scholar during our data collection. We recommend that authors make sure that, when preprints end up getting published, they update their preprint with the publication DOI when it is available. Particularly if their title has changed. This is quite easy for authors to do in *EcoEvoRxiv* and it will ensure that the preprint is linked to the published article and the citations are merged. It will also create a link between the final preprint and the published paper so that, no matter the open access status of the publication, the reserach findings will remain openly accessible, saving costs for authors and improving the visibility and use of research.

Despite the early successes of the new initiatives taken by *EcoEvoRxiv* it is still clear that much work needs to be done to improve our communities understanding of preprints (and postprints). We view this perspective piece as one small step in achieving that goal. Authors also need to be aware of the benefits of using community driven preprint servers and the unique initiatives being pursued. Community driven preprint servers are capable of setting their own agenda and are driven by the needs and disires of the community they would like their findings to be shared with. Supporting these initiatives should be a priority for all researchers and those volunteering their time at *EcoEvoRxiv* remain open to new and innovative ways to improve publication and open science practices.  We believe that the future of preprints is bright and community driven initiatives, such as *EcoEvoRxiv*, are set to play a key role in the future of scientific publishing.
